---
title: "Lesson 3 - Additional Visualizations"
author: "Caitlin Eger"
date: "August 16, 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Making pretty pictures

Because R is an open source software package, there are many different packages for generating graphs, maps, animations, etc.  We won't be able to introduce everything during this workshop, but we want to give you enough tools to get started.  Remember, a quick internet search usually turns up some helpful examples if you are struggling to get your plot to look right.

### Base R images

The functions included with your basic R install are actually super flexible.  It is possible to make a figure do pretty much anything using the base R plotting functions.  The downside of all this flexibility is you might need to do a lot of coding to get your plot perfect.

Lets look at some examples using some data from Lesson 1

```{r}
FLP2018 <- readRDS("data/FLP2018.Rdata")

plot(FLP2018$Site, FLP2018$Temp, 
     xlab = "Site",
     ylab = "Temperature in degrees C")
```

That was pretty quick and painless!  It also only scratches the surface of what these plots can do, but it can take a lot of work to learn all the functions you need.

Another example
```{r}
par(mfrow = c(1,2))

plot(FLP2018$Site, FLP2018$Temp, 
     xlab = "Site",
     ylab = "Temperature in degrees C")

plot(FLP2018$Depth, FLP2018$Temp, col = FLP2018$Date,
     xlab = "Depth",
     ylab = "Temperature in degrees C")
```

The second figure here is a depth profile, but it doesn't look like one.

```{r}
par(mfrow = c(1,2))

plot(FLP2018$Site, FLP2018$Temp, 
     xlab = "Site",
     ylab = "Temperature in degrees C")

plot(y=FLP2018$Depth, x=FLP2018$Temp, col = FLP2018$Date,
     ylab = "Depth",
     xlab = "Temperature in degrees C",
    ylim=rev(range(FLP2018$Depth)))
```

Also notice the order in which you run lines of code is important when using the basic R graphics.  If we want to put a line, or additional points on the plot, we need to make sure everything happens in the right order.

### ggplot
